---
title: "Module 3: Data Structures in R"
toc: true
---

## Introduction

In the previous module, we learned that R stores information in **objects**.

So far, most of our objects contained only one value:

```{r}
number_of_cameras <- 12

species <- "Raccoon"

camera_active <- TRUE
```

But real datasets contain much more information than a single number or word.

R allows us to organize multiple values into different **data structures**.

In this module, we will focus on four of the most common:

- vectors,
- matrices,
- lists,
- data frames.

Understanding the difference between these structures will make it much easier to understand what R is doing later when we start working with real datasets.

## Let's get started

Open your `intro-r-course` project and create a new R script.

Save it inside the `scripts` folder as:

```text
module-03.R
```

We will use this script throughout the module.

## Vectors

A **vector** is one of the most basic and important data structures in R.

A vector allows us to store several values inside a single object.

We can create a vector using the function:

```text
c()
```

The `c` stands for **combine**.

For example:

```{r}
detections <- c(5, 12, 8, 0, 3)

detections
```

Instead of creating five different objects:

```{r}
detection_1 <- 5
detection_2 <- 12
detection_3 <- 8
detection_4 <- 0
detection_5 <- 3
```

we can store all five values together:

```{r}
detections <- c(5, 12, 8, 0, 3)
```

This makes it much easier to work with several related observations.

### Numeric vectors

Vectors can contain numeric values:

```{r}
camera_days <- c(30, 28, 30, 25, 29)

camera_days
```

We can perform operations directly on the entire vector:

```{r}
camera_days + 1
```

or:

```{r}
camera_days * 2
```

R applies the operation to **each element of the vector**.

We can also use functions:

```{r}
mean(camera_days)
```

```{r}
sum(camera_days)
```

```{r}
length(camera_days)
```

The function `length()` tells us how many elements are contained in the vector.

### Character vectors

Vectors can also contain text:

```{r}
species <- c(
  "Raccoon",
  "Coyote",
  "Bobcat",
  "White-tailed deer"
)

species
```

And logical values:

```{r}
camera_active <- c(TRUE, TRUE, FALSE, TRUE)

camera_active
```

::: {.callout-important}
## Vectors contain one type of data

An atomic vector can contain only **one basic data type**.

For example, all its values can be numeric, all can be character, or all can be logical.
:::

What happens if we mix types?

```{r}
mixed_vector <- c(1, 2, "Raccoon", 4)

mixed_vector
```

R converts all the values to a compatible type.

Check:

```{r}
class(mixed_vector)
```


Because the vector contains text, R converted the numbers into character values too.

This automatic conversion is called **coercion**.

::: {.callout-tip}
If an object that you expected to be numeric suddenly becomes `character`, check whether one of its values contains text.
:::

## Creating sequences

Sometimes we do not want to type every number individually.

For example:

```{r}
1:10
```


We can save it as an object:

```{r}
sampling_days <- 1:10

sampling_days
```

The function `seq()` gives us more control. This function has several arguments: `from` is the starting number, `to` is the ending number, `length.out` is used to control the length of the vector, and `by` allows us to specify the interval of the sequence.

```{r}
seq(from = 0, to = 30, by = 5)
```


We can also repeat values using `rep()`. The `times` argument allows us to specify the number of times a value or vector will be repeated. When we specify `each`, we can control the number of times each value within the vector is repeated.

```{r}
rep("Camera", times = 5)
```

These functions are useful when we need to generate repeated or sequential values without typing everything manually.

## Selecting elements from a vector

Sometimes we only want one or a few values from a vector.

We can select elements using square brackets:

```text
[]
```

Let's create a vector:

```{r}
detections <- c(5, 12, 8, 0, 3)
```

To select the first value:

```{r}
detections[1]
```

To select the third:

```{r}
detections[3]
```

We can select several positions:

```{r}
detections[c(1, 3, 5)]
```

Or a sequence of positions:

```{r}
detections[1:3]
```

::: {.callout-important}
R starts counting at **1**, not at 0.

The first element of a vector is:

```{r}
#| eval: false
#| 
x[1]
```

:::

We can also exclude positions using negative numbers:

```{r}
detections[-1]
```

This returns everything except the first element.

### Selecting values using a condition

Remember the relational operators from the previous module?

We can use them with vectors.

For example:

```{r}
detections > 5
```

R returns a logical value for every element:



We can use this condition inside the brackets:

```{r}
detections[detections > 5]
```

R returns only the values that meet the condition:


This idea will become extremely important later when we start filtering datasets.

::: {.callout-note}
## Try it yourself

Create this vector:

```{r}
camera_days <- c(30, 21, 29, 15, 30, 27)
```

Then:

1. Select the second value.
2. Select the first three values.
3. Select values 2, 4, and 6.
4. Find which values are greater than 25.
5. Return only the values greater than 25.
:::

## Matrices

A **matrix** is a two-dimensional data structure.

Unlike a vector, which has only one dimension, a matrix has:

- rows,
- columns.

We can create a matrix using the function `matrix()`.

For example:

```{r}
camera_matrix <- matrix(
  1:12,
  nrow = 4,
  ncol = 3
)

camera_matrix
```

R fills matrices by column by default.

We can change this using:

```{r}
camera_matrix <- matrix(
  1:12,
  nrow = 4,
  ncol = 3,
  byrow = TRUE
)

camera_matrix
```

Now the values are filled by row.

### Matrix dimensions

We can inspect the dimensions of a matrix using:

```{r}
dim(camera_matrix)
```

This returns the number of rows and columns.

We can also use:

```{r}
nrow(camera_matrix)
```

and:

```{r}
ncol(camera_matrix)
```

### Creating matrices from vectors

We can combine vectors into a matrix.

For example:

```{r}
station_1 <- c(5, 2, 0)
station_2 <- c(3, 1, 4)
station_3 <- c(8, 0, 2)
```

We can combine them by rows:

```{r}
detection_matrix <- rbind(
  station_1,
  station_2,
  station_3
)

detection_matrix
```

Or by columns:

```{r}
cbind(
  station_1,
  station_2,
  station_3
)
```

The functions are easy to remember:

```text
rbind = row bind

cbind = column bind
```

### Naming rows and columns

Matrices can have row and column names.

For example:

```{r}
rownames(detection_matrix) <- c(
  "CT01",
  "CT02",
  "CT03"
)

colnames(detection_matrix) <- c(
  "Raccoon",
  "Coyote",
  "Bobcat"
)

detection_matrix
```

Now our matrix is much easier to interpret.

This type of structure will become familiar later when we work with camera-trap detection histories.

### Selecting elements from a matrix

Because matrices have two dimensions, we need to specify:

```text
[row, column]
```

For example:

```{r}
detection_matrix[1, 2]
```

means:

> row 1, column 2

We can select an entire row:

```{r}
detection_matrix[1, ]
```

Or an entire column:

```{r}
detection_matrix[, 2]
```

We can also use names:

```{r}
detection_matrix["CT01", "Coyote"]
```

or:

```{r}
detection_matrix[, "Raccoon"]
```

::: {.callout-tip}
A useful way to remember matrix indexing is:

```text
object[row, column]
```

The comma separates the two dimensions.
:::

We can also select values from the diagonal using `diag`

```{r}
diag(detection_matrix)
```

Delete row 1 and select column 2
```{r}

detection_matrix[-1,2]
```



Delete rows 1 and 4, and select columns 1 and 3
```{r}
detection_matrix[c(-1,-4), c(1,3)]
```

When we use the selectors on the left side of the object definition, we can replace the values with the ones we are defining.

```{r}
# We can replace values in matrices

# Replace the second row with 100, 200, and 300
detection_matrix[2,] <- c(100,200, 300)
detection_matrix
```

### Matrices contain one type of data

Like vectors, matrices contain only one basic type of data.

For example:

```{r}
numeric_matrix <- matrix(1:9, nrow = 3)

numeric_matrix
```

But if we introduce text:

```{r}
mixed_matrix <- cbind(
  station = c("CT01", "CT02", "CT03"),
  detections = c(5, 8, 2)
)

mixed_matrix
```

R converts the numeric values into characters because the entire matrix must have a common type.

This limitation is one of the reasons why **data frames** are more useful for most datasets.

## Lists

A **list** is a very flexible R data structure.

Unlike vectors and matrices, lists can contain objects of **different types and structures**.

For example:

```{r}
camera_information <- list(
  station = "CT01",
  active = TRUE,
  detections = 25,
  species = c("Raccoon", "Coyote", "Bobcat")
)

camera_information
```

This single list contains:

- a character value,
- a logical value,
- a numeric value,
- a character vector.

A list can even contain matrices, data frames, or other lists.

For example:

```{r}
example_list <- list(
  numbers = c(1, 2, 3),
  matrix = matrix(1:4, nrow = 2),
  message = "Hello R"
)

example_list
```

This flexibility makes lists extremely common in R.

Many functions and statistical models return their results as lists containing several different objects.

### Selecting elements from a list

If the elements have names, we can use `$`:

```{r}
camera_information$station
```

```{r}
camera_information$species
```

We can also use double brackets:

```{r}
camera_information[[1]]
```

or:

```{r}
camera_information[["station"]]
```

You may also see:

```{r}
camera_information[1]
```

but there is an important difference.

returns a **list containing the first element**.

In contrast:

```{r}
camera_information[[1]]
```

returns the **contents of the first element**.

This distinction can be confusing at first, and you do not need to memorize it yet.

For now, the most important idea is:

> Lists can store many different kinds of objects together.

## Data frames

For the type of work we will do in this course, **data frames will probably be the most important data structure**.

A data frame is a rectangular table with:

- rows,
- columns.

It looks very similar to an Excel spreadsheet.

However, unlike a matrix, **different columns of a data frame can contain different types of data**.

For example, let's create some information from a hypothetical camera-trap survey:

```{r}
station <- c(
  "CT01",
  "CT02",
  "CT03",
  "CT04"
)

habitat <- c(
  "Forest",
  "Forest",
  "Pasture",
  "Pasture"
)

camera_days <- c(
  30,
  28,
  30,
  27
)

active <- c(
  TRUE,
  TRUE,
  FALSE,
  TRUE
)
```

Each object is a vector.

Now we can combine them into a data frame:

```{r}
camera_data <- data.frame(
  station,
  habitat,
  camera_days,
  active
)

camera_data
```

We now have one object containing several columns.

Each column can have a different type:

```{r}
class(camera_data$station)
```

```{r}
class(camera_data$camera_days)
```

```{r}
class(camera_data$active)
```

The data frame itself has its own class:

```{r}
class(camera_data)
```

R returns:


::: {.callout-important}
## Rows and columns usually represent different things

In most ecological datasets:

- **rows** represent observations,
- **columns** represent variables.

Understanding this organization is extremely important when preparing data for analysis.
:::

In our example:

```text
station      habitat      camera_days      active
```

are variables.

Each row represents one camera station.

## Inspecting a data frame

Before analyzing a dataset, we should always look at its structure.

Some useful functions are:

```{r}
camera_data
```

### `head()`

Shows the first few rows:

```{r}
head(camera_data)
```

### `str()`

Shows the internal structure of the object:
```{r}
str(camera_data)
```

### `dim()`

Returns the number of rows and columns:

```{r}
dim(camera_data)
```

### `nrow()`

Returns the number of rows:

```{r}
nrow(camera_data)
```

### `ncol()`

Returns the number of columns:
```{r}
ncol(camera_data)
```

### `names()`

Returns the column names:
```{r}
names(camera_data)
```



::: {.callout-tip}
When you load a new dataset into R, one of the first things you should do is inspect it.

Functions such as:

```r
head()
str()
dim()
names()
```

can help you identify problems before you begin an analysis.
:::

## Selecting columns from a data frame

We can access a column using `$`:

```{r}
camera_data$station
```

```{r}
camera_data$camera_days
```

Notice what happens when we do this:

```{r}
class(camera_data$camera_days)
```

The column itself is a vector.

This is an important idea:

> A data frame is essentially a collection of equal-length vectors organized as columns.

We can also use square brackets.

For example:

```{r}
camera_data[1, 2]
```

means:

> row 1, column 2

To select the first row:

```{r}
camera_data[1, ]
```

To select the second column:

```{r}
camera_data[, 2]
```

We can also use column names:

```{r}
camera_data[, "habitat"]
```

Or select several columns:

```{r}
camera_data[, c("station", "camera_days")]
```

Later we will learn easier and more readable ways to select rows and columns using `dplyr`.

For now, it is important to understand how data frames work in base R.

## Adding a column

We can add a new column using `$`.

For example:

```{r}
camera_data$elevation <- c(
  105,
  112,
  98,
  101
)

camera_data
```

The new column must have the correct number of values.

Our data frame has four rows, so the new column needs four values.

We can verify the number of rows with:

```{r}
nrow(camera_data)
```

## Changing values

We can also replace specific values.

For example:

```{r}
camera_data$camera_days[2] <- 30

camera_data
```

Or using row and column positions:

```{r}
camera_data[2, "camera_days"] <- 30
```

This can be useful, but we should be careful when modifying raw data manually.

::: {.callout-important}
Whenever possible, avoid manually changing your original data file.

It is usually better to make corrections through code so that the changes are documented and reproducible.
:::

## Different structures, different purposes

Let's summarize the four structures we have seen.

| Structure | Dimensions | Same data type? | Common use |
|---|---:|---|---|
| Vector | 1 | Yes | Store a sequence of values |
| Matrix | 2 | Yes | Numeric or structured rectangular data |
| List | 1 | No | Store different objects together |
| Data frame | 2 | Different types allowed by column | Store datasets |

The structure we choose depends on the type of information we want to store.

Throughout the rest of the course:

- we will constantly work with **vectors**,
- most of our datasets will be **data frames**,
- we will occasionally encounter **matrices**,
- and many R functions will return results as **lists**.

When we eventually work with `camtrapR`, all four of these ideas will appear again.

## Converting between data structures

R also provides functions for converting objects from one structure to another.

For example:

```{r}
as.data.frame(detection_matrix)
```

converts a matrix into a data frame.

We can save the result:

```{r}
detection_data <- as.data.frame(detection_matrix)

class(detection_data)
```

Similarly:

```{r}
as.matrix(camera_data)
```

attempts to convert a data frame into a matrix.

However, remember that a matrix can contain only one basic data type.

Because `camera_data` contains text, numbers, and logical values, R will need to convert everything to a common type.

Try:

```{r}
camera_matrix_2 <- as.matrix(camera_data)

camera_matrix_2
```

and then:

```{r}
class(camera_matrix_2)
```

This is another good example of why understanding data structures matters.

A function may technically convert an object, but the result may not have the structure or data type that you expected.

## Practice at home

::: {.callout-tip}
This activity should take approximately **15–20 minutes**.

Open your `intro-r-course` project and create a new script called:

```text
practice-03.R
```

Save it inside the `scripts` folder.

### Part 1: Vectors

Create these objects:

```{r}
stations <- c("CT01", "CT02", "CT03", "CT04", "CT05")

detections <- c(12, 5, 0, 18, 7)

camera_days <- c(30, 28, 30, 25, 29)
```

Then:

1. Return the third element of `stations`.
2. Return the first three values of `detections`.
3. Determine which values of `detections` are greater than 10.
4. Return only the detections greater than 10.
5. Calculate the mean number of camera-days.

### Part 2: Data frame

Create a data frame called `survey_data` using:

```{r}
stations
detections
camera_days
```

Then:

1. Display the data frame.
2. Use `str()` to inspect its structure.
3. Use `dim()` to find its dimensions.
4. Use `names()` to obtain its column names.
5. Select only the `detections` column.
6. Select the information from the second camera station.
7. Add a new column called `habitat` containing:

```{r}
c("Forest", "Forest", "Pasture", "Forest", "Pasture")
```

8. Use `str()` again and look at the type of each column.

:::

::: {.callout-tip collapse="true"}
## Suggested solution

```{r}
# Practice for Module 3

# ---------------------------
# Part 1: Vectors
# ---------------------------

stations <- c(
  "CT01",
  "CT02",
  "CT03",
  "CT04",
  "CT05"
)

detections <- c(
  12,
  5,
  0,
  18,
  7
)

camera_days <- c(
  30,
  28,
  30,
  25,
  29
)

# Third station
stations[3]

# First three detection values
detections[1:3]

# Which values are greater than 10?
detections > 10

# Return only detections greater than 10
detections[detections > 10]

# Mean camera-days
mean(camera_days)


# ---------------------------
# Part 2: Data frame
# ---------------------------

survey_data <- data.frame(
  stations,
  detections,
  camera_days
)

# Display the data
survey_data

# Inspect structure
str(survey_data)

# Dimensions
dim(survey_data)

# Column names
names(survey_data)

# Select detections
survey_data$detections

# Select the second camera station
survey_data[2, ]

# Add habitat
survey_data$habitat <- c(
  "Forest",
  "Forest",
  "Pasture",
  "Forest",
  "Pasture"
)

# Inspect the updated data frame
str(survey_data)
```

The important thing is that you understand the difference between:

- a vector,
- a matrix,
- a list,
- and a data frame,

and that you can recognize how rows, columns, and positions are used to access information.
:::