---
title: "Module 4: Importing and Exporting Data"
toc: true
---


```{r}
#| echo: false
#| output: false
library(here)
```


## Introduction

So far, we have created all of our objects directly inside R.

For example:

```{r}
stations <- c("CT01", "CT02", "CT03")

detections <- c(12, 5, 8)
```

But this is not how we normally work with real datasets.

Most of the time, our data already exist in a file created in Excel, Google Sheets, field-data software, or another program.

One of the most important skills when learning R is therefore knowing how to **bring external data into R**.

In this module we will learn how to:

- understand where R looks for files,
- organize data inside an RStudio Project,
- import `.csv` files,
- import Excel files,
- inspect imported data,
- and export data from R.


::: {.callout-note}
## Download the example data

We will use these files throughout this module.

[Download CSV](../data/camera_data.csv){.btn .btn-primary download="camera_data.csv"}

[Download Excel](../data/camera_data.xlsx){.btn .btn-secondary download="camera_data.xlsx"}
:::

## Let's get started

Open your `intro-r-course` project and create a new R script.

Save it inside the `scripts` folder as:

```text
module-04.R
```

Remember that our project should look approximately like this:

```text
intro-r-course/
│
├── data/
├── scripts/
│   └── module-04.R
├── outputs/
└── intro-r-course.Rproj
```

Throughout this module, the files we want to import will be stored inside:

```text
data/
```

## Where is R looking for files?

Before importing a file, R needs to know **where the file is located**.

The folder that R currently considers its starting location is called the **working directory**.

We can check it using:

```{r}
getwd()
```

Because we are working inside an RStudio Project, the working directory should normally be the main folder of our project:

```text
intro-r-course/
```

This is one of the main reasons we created an RStudio Project at the beginning of the course.

::: {.callout-important}
## Avoid using `setwd()` inside your scripts

You may see tutorials that use:

```r
setwd("C:/Users/YourName/Documents/MyProject")
```

to specify the working directory.

This may work on your computer, but that path will probably not work on someone else's computer.

Using an RStudio Project allows us to avoid this problem and work with **relative paths** instead.
:::

## File paths

A **file path** tells R where a file is located.

There are two general types of paths:

- absolute paths,
- relative paths.

### Absolute paths

An absolute path describes the complete location of a file on your computer.

For example:

```text
C:/Users/Gabriel/Documents/intro-r-course/data/camera_data.csv
```

This may work perfectly on one computer.

But if we send the project to another person, their computer probably does not have:

```text
C:/Users/Gabriel/
```

So the code will no longer work.

### Relative paths

A relative path describes the location of a file **relative to our project folder**.

If our file is stored here:

```text
intro-r-course/
│
├── data/
│   └── camera_data.csv
│
├── scripts/
└── outputs/
```

we can refer to it simply as:

```text
data/camera_data.csv
```

This is much more portable.

If we move the entire project to another computer, the internal structure remains the same.

::: {.callout-tip}
## Keep your project organized

A simple project structure such as:

```text
project/
├── data/
├── scripts/
└── outputs/
```

makes file paths much easier to understand and helps keep your analyses reproducible.
:::

## What is a CSV file?

One of the most common formats for storing tabular data is **CSV**.

CSV stands for:

> **Comma-Separated Values**

A CSV file is essentially a plain-text table.

For example, a spreadsheet that looks like this:

| station | species | detections | camera_days |
|---|---|---:|---:|
| CT01 | Raccoon | 12 | 30 |
| CT02 | Coyote | 5 | 28 |
| CT03 | Bobcat | 8 | 30 |

may be stored internally as:

```text
station,species,detections,camera_days
CT01,Raccoon,12,30
CT02,Coyote,5,28
CT03,Bobcat,8,30
```

Because CSV files are simple text files, they can be opened by many different programs, including:

- Excel,
- Google Sheets,
- R,
- Python,
- text editors,
- and many other programs.

This makes CSV a very useful format for sharing and analyzing data.

## From Excel to CSV

Many datasets begin their life in Excel.

That is completely fine.

However, when we begin analyzing the data in R, it is often useful to save a copy as a CSV file.

In Excel:

**File → Save As**

and select:

```text
CSV (Comma delimited) (*.csv)
```

Save the file inside the `data` folder of your RStudio Project.

For example:

```text
data/camera_data.csv
```

::: {.callout-important}
## Keep your original data

Do not replace or modify your only copy of the original dataset.

A good workflow is to keep the original file unchanged and create a separate CSV file for analysis.

For example:

```text
data/
├── camera_data_original.xlsx
└── camera_data.csv
```
:::

## Importing a CSV file

Base R includes the function:

```text
read.csv()
```

which reads a CSV file and converts it into a data frame.

If our file is located here:

```text
data/camera_data.csv
```

we can import it using:



```{r}
#| eval: false

camera_data <- read.csv("data/camera_data.csv")
```

```{r}
#| include: false
camera_data <- read.csv(here("data", "camera_data.csv"))
```


There are two important things happening here.

First:

```r
read.csv("data/camera_data.csv")
```

reads the file.

Second:

```r
camera_data <-
```

stores the imported dataset in an R object called `camera_data`.

If we forget the assignment:




```{r}
#| eval: false

read.csv("data/camera_data.csv")
```

```{r}
#| echo: false
#| 
read.csv(here("data", "camera_data.csv"))
```

R can still read and display the file, but we have not saved it as an object that we can continue working with.

## Check that the data loaded correctly

Importing a file without an error does not necessarily mean that everything was read correctly.

We should always inspect the data after importing it.

For example:

```{r}


head(camera_data)
```

```{r}


str(camera_data)
```

```{r}


dim(camera_data)
```

```{r}


names(camera_data)
```

These functions should already look familiar from the previous module.

We can also open the dataset using RStudio's data viewer:

```{r}
#| eval: false

View(camera_data)
```

::: {.callout-tip}
## Always check your imported data

After importing a dataset, ask yourself:

- Does it have the expected number of rows?
- Does it have the expected number of columns?
- Are the column names correct?
- Are numeric columns actually numeric?
- Did missing values import correctly?
- Does the dataset look like the file I expected to load?

Never assume that a file was imported correctly just because R did not return an error.
:::

## Common CSV problems

Files do not always import exactly as expected.

A few common problems include:

### Wrong file name

For example:

```{r}
#| eval: false

read.csv("data/camera-data.csv")
```

will not work if the actual file is called:

```text
camera_data.csv
```

The names must match.

### Wrong folder

This:

```{r}
#| eval: false

read.csv("camera_data.csv")
```

tells R to look for the file directly inside the project folder.

But if the file is inside:

```text
data/
```

we need:

```{r}
#| eval: false

read.csv("data/camera_data.csv")
```

### Capitalization

Depending on the operating system:

```text
Camera_Data.csv
```

and:

```text
camera_data.csv
```

may be treated as different file names.

Consistent file names help avoid these problems.

## Does the file exist?

If R tells you that it cannot find a file, we can check whether R can see it.

For example:

```{r}
#| eval: false

file.exists("data/camera_data.csv")
```

```{r}
#| echo: false
file.exists(here("data", "camera_data.csv"))
```


If R returns:

```text
TRUE
```

the file exists at that location.

If R returns:

```text
FALSE
```

something about the path or file name is incorrect.

We can also see the files inside our `data` folder:

```{r}
#| eval: false

list.files("data")
```

```{r}
#| echo: false
list.files(here("data"))
```


These two functions can be extremely useful when troubleshooting import problems.

## Importing CSV files with `readr`

There are several ways to import the same type of file in R.

The `tidyverse` includes a package called `readr` that provides another function for reading CSV files:

```text
read_csv()
```

Because we installed `tidyverse` at the beginning of the course, we can load it using:

```{r}

library(tidyverse)
```

Then:

```{r}
#| eval: false

camera_data2 <- read_csv("data/camera_data.csv")
```

```{r}
#| include: false
camera_data2 <- read_csv(here("data", "camera_data.csv"))
```



Notice the difference:

```text
Base R       read.csv()

readr        read_csv()
```

Both functions can read CSV files.

For this course, you may see both, but later we will frequently use `read_csv()` because it works naturally with other tidyverse tools.

::: {.callout-note}
You do not need to memorize multiple ways to import the same file.

What matters is understanding:

1. where the file is,
2. what type of file it is,
3. which function can read it,
4. and what object will store the result.
:::

## Importing Excel files

Sometimes we want to import an Excel file directly instead of converting it to CSV.

Base R does not include a function for reading `.xlsx` files, so we need an additional package.

We installed `readxl` during the Getting Started section.

Load it using:

```{r}

library(readxl)
```

Then we can import an Excel file using:

```{r}
#| eval: false

camera_data3 <- read_excel( "data/camera_data.xlsx")
```

```{r}
#| include: false
camera_data3 <- read_excel(here("data", "camera_data.xlsx"))
```


The function is called:

```text
read_excel()
```

and comes from the `readxl` package.

### Excel files with multiple sheets

An Excel workbook can contain several sheets.

We can see their names using:

```{r}
#| eval: false

excel_sheets("data/camera_data.xlsx")
```

```{r}
#| echo: false

excel_sheets(here("data", "camera_data.xlsx"))
```


If the workbook contains a sheet called:

```text
camera_records
```

we can import that specific sheet using:

```{r}
#| eval: false

camera_data3 <- read_excel("data/camera_data.xlsx",
  sheet = "camera_records"
)
```

```{r}
#| echo: false

camera_data3 <- read_excel( here("data","camera_data.xlsx"), sheet = "camera_records")

```


We can also select a sheet by position:

```{r}
#| eval: false

camera_data3 <- read_excel(
  "data/camera_data.xlsx",
  sheet = 1
)
```

## CSV or Excel?

Both formats can be useful.

### CSV

Advantages:

- simple,
- open format,
- easy to share,
- readable by almost any software,
- good for analysis and reproducibility.

Limitations:

- only stores one table,
- does not preserve formatting,
- does not store formulas or multiple sheets.

### Excel

Advantages:

- familiar to many users,
- can contain multiple sheets,
- useful for entering and reviewing data.

Limitations:

- more complex file format,
- formatting can sometimes hide data problems,
- spreadsheet software may automatically modify dates or other values.

A common workflow is:

> **Enter/review data in Excel → save a clean CSV → analyze the CSV in R**

But R can work directly with either format.

## Importing data using RStudio

RStudio also provides a graphical interface for importing datasets.

In the **Environment** panel, select:

**Import Dataset**

RStudio will provide options for importing text and Excel files.

This can be useful when you are learning because RStudio shows a preview of the data and generates the corresponding R code.

However, once you understand the import function, it is better to keep the import command inside your script.

For example:

```{r}
#| eval: false

camera_data <- read_csv(
  "data/camera_data.csv"
)
```

This way, the analysis can be reproduced later without manually clicking through menus.

::: {.callout-tip}
The graphical import tool is useful for learning.

But try to copy the generated code into your script so that you can reproduce the import later.
:::

## File names and column names

Good file organization begins before we import anything into R.

Try to use simple and consistent file names.

Instead of:

```text
Camera Data FINAL version 2 updated!!!.xlsx
```

use something like:

```text
camera_data.xlsx
```

or:

```text
camera_data_2026.xlsx
```

The same principle applies to column names.

Instead of:

```text
Camera Station
Number of detections
Habitat Type
```

consider:

```text
station
detections
habitat
```

or:

```text
camera_station
number_of_detections
habitat_type
```

Consistent names make working with data in R much easier.

## Exporting data

Sometimes we want to save a dataset after modifying or summarizing it in R.

For CSV files, base R provides:

```text
write.csv()
```

For example:

```{r}
#| eval: false

write.csv(
  camera_data,
  "outputs/camera_data_clean.csv",
  row.names = FALSE
)
```

This creates:

```text
outputs/camera_data_clean.csv
```

We use:

```text
row.names = FALSE
```

because otherwise base R will add an extra column containing the row numbers.

### Exporting with `readr`

If we are using the tidyverse, we can also use:

```{r}
#| eval: false

write_csv(
  camera_data,
  "outputs/camera_data_clean.csv"
)
```

`write_csv()` does not add row names by default.

Again, you do not need to memorize every alternative.

The important idea is:

```text
read  → bring data into R

write → save data from R
```

## Keep raw data separate from outputs

Remember our project structure:

```text
intro-r-course/
│
├── data/
├── scripts/
└── outputs/
```

A useful rule is:

```text
data/
```

contains the original data that we received or collected.

```text
outputs/
```

contains files created by our analysis.

For example:

```text
intro-r-course/
│
├── data/
│   └── camera_data.csv
│
├── scripts/
│   └── module-04.R
│
└── outputs/
    └── camera_data_clean.csv
```

::: {.callout-important}
Try not to overwrite your original data from R.

Keeping the raw data unchanged means that you can always return to the original information and reproduce every change using your script.
:::

## Practice at home

::: {.callout-tip}
This activity should take approximately **15–20 minutes**.

Open your `intro-r-course` project and create a new script called:

```text
practice-04.R
```

Save it inside the `scripts` folder.

Use any small Excel dataset you have available, or create one with at least:

- 5 rows,
- 3 columns,
- one column containing text,
- one column containing numbers.

For example:

| station | habitat | detections |
|---|---|---:|
| CT01 | Forest | 8 |
| CT02 | Pasture | 3 |
| CT03 | Forest | 12 |
| CT04 | Pasture | 0 |
| CT05 | Forest | 5 |

### Part 1: Prepare the data

1. Save the Excel file inside your project's `data` folder.
2. Save a second copy as a `.csv` file.

Your folder should contain something like:

```text
data/
├── practice_data.xlsx
└── practice_data.csv
```

### Part 2: Import the CSV

Import the CSV file into an object called:

```text
practice_data
```

Then use:

```r
head()
str()
dim()
names()
```

to inspect the imported dataset.

### Part 3: Import the Excel file

Load the `readxl` package and import the Excel version of the same dataset.

Save it as:

```text
practice_excel
```

Compare:

```r
class(practice_data)
class(practice_excel)
```

and inspect both objects.

### Part 4: Export the data

Export `practice_data` into the `outputs` folder with the name:

```text
practice_data_exported.csv
```

Finally, check that the new file appears inside the `outputs` folder.

:::

::: {.callout-tip collapse="true"}
## Suggested solution

```{r}
#| eval: false

# Practice for Module 4

# ---------------------------
# Import CSV
# ---------------------------

practice_data <- read.csv(
  "data/practice_data.csv"
)

# Inspect the data
head(practice_data)
str(practice_data)
dim(practice_data)
names(practice_data)


# ---------------------------
# Import Excel
# ---------------------------

library(readxl)

practice_excel <- read_excel(
  "data/practice_data.xlsx"
)

# Inspect the Excel data
head(practice_excel)
str(practice_excel)

# Compare classes
class(practice_data)
class(practice_excel)


# ---------------------------
# Export data
# ---------------------------

write.csv(
  practice_data,
  "outputs/practice_data_exported.csv",
  row.names = FALSE
)
```

The important thing is that you understand the complete workflow:

```text
external file
      ↓
relative path
      ↓
read the file
      ↓
store it as an R object
      ↓
inspect the object
      ↓
work with the data
      ↓
write an output file
```

:::